---
title: "Batch correction"
author: "Katerina Danko"
output:
    html_document:
      code_folding: show
      theme: united
      toc: true
      toc_depth: 3
      toc_float: true
      number_section: true
---

```{css, echo=FALSE}
.header-section-number::after {
  content: ".";
}
```

List of used packages:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r include=FALSE}
# "data.Rdata" may be found in "Rdata" directory 
load("data.Rdata")
```

```{r echo = FALSE, message=FALSE, warning=FALSE}
# Missing packages must be installed for proper compilation 
if (!require(limma)) BiocManager::install("limma")
if (!require(dplyr)) install.packages("dplyr")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(bapred)) install.packages("bapred")
if (!require(Harman)) BiocManager::install("Harman")
if (!require(sva)) BiocManager::install("sva")
if (!require(devtools)) install.packages("devtools")
if (!require(ggbiplot)) devtools::install_github("vqv/ggbiplot")
if (!require(heatmaply)) BiocManager::install("heatmaply")
if (!require(ggVennDiagram)) devtools::install_github("gaospecial/ggVennDiagram")
if (!require(EnhancedVolcano)) BiocManager::install("EnhancedVolcano")
if (!require(mixOmics)) BiocManager::install("mixOmics")
```

```{r echo = T, warning=FALSE, message=FALSE, eval=TRUE}
library(limma)
library(bapred)
library(Harman)
library(ggplot2)
library(sva)
library(ggbiplot)
library(heatmaply)
library(EnhancedVolcano)
library(ggVennDiagram)
library(mixOmics)
library(dplyr)
```

```{r echo = FALSE, warning=FALSE, message=FALSE, eval=TRUE}
theme_set(theme_bw())
```

# BMC
Limma package provides *BMC (batch mean centering)* method for batch correction.
BMC method includes centering the variables within batches to have zero mean. 

## Batch correction
```{r warning=FALSE, message=FALSE}
data_limma <- data_norm_quantile_max
batch <- as.numeric(data_factors$Series)

ex <- cbind(data_limma, batch)

data_limma_corrected <- removeBatchEffect(data_limma, batch)
```

## PCA
```{r warning=FALSE, message=FALSE}
# Year 1 and 2
pca_limma <- prcomp(t(data_limma_corrected), center = T, scale. = F)

ggbiplot(pca_limma, ellipse = TRUE, groups = data_factors$Series, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  labs(title = "After correction (BMC)", color = "Year") +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

ggbiplot(pca_limma, ellipse = TRUE, groups = data_factors$Differentiation, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  geom_point(aes(col = data_factors$Differentiation, shape = data_factors$Series), size = 2) +
  labs(title = "After correction (BMC)", color = "Group", shape = "Year") +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

## Differential expression
```{r warning=FALSE, message=FALSE}
X_year <- model.matrix(~ Series, data = data_factors)

fit_y <- lmFit(data_limma_corrected, design = X_year, method = "robust", maxit = 1000)

efit_y <- eBayes(fit_y)

topTable(efit_y, coef = 2)

num_spots <- nrow(data_norm_quantile_max)
full_list_y <- topTable(efit_y, coef = 2, number = num_spots,
                        sort.by = "none")
```

Draw heatmap:
```{r warning=FALSE, message=FALSE}
p_above_y <- full_list_y$adj.P.Val <= 0.05
dif_data_limma <- data_norm_quantile_max[p_above_y, ]

heatmaply(dif_data_limma, main = "Year 1 and 2", fontsize_row = 1, k_col = NA, dendrogram = "col", scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(low = "lightseagreen", high = "orangered3", midpoint = 15))
```

Draw Volcano plot: 
```{r warning=FALSE, message=FALSE}
EnhancedVolcano(full_list_y,
                lab = rownames(full_list_y),
                title = "After correction (BMC)",
                subtitle = NULL,
                x = 'logFC',
                y = 'adj.P.Val',
                pCutoff = 0.05,
                FCcutoff = 0.1,
                col = c("lightcyan4","#f57002", "#ee9f02", "#0a9278"),
                legend = c("Not significant","Log2FC","Padj","Padj & Log2FC"),
                legendPosition = "right")
```

# Ratio A
*Ratio A* method of batch correction is ratio-based method scaling the expression values by the arithmetic mean. "bapred" package provides possibility to correct batch effect by Ratio A method. 

## Batch correction:
```{r warning=FALSE, message=FALSE}
batch <- data_factors$Series
ratio_A_data <- ratioa(t(data_norm_quantile_max), batch)
ratio_A_corrected <- ratio_A_data$xadj
```
## PCA
```{r warning=FALSE, message=FALSE}
pca_ratio_A <- prcomp(ratio_A_corrected, center = T, scale. = F)

ggbiplot(pca_ratio_A, ellipse = TRUE, groups = data_factors$Series, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "After correction (Ratio A)", color = "Year") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

ggbiplot(pca_ratio_A, ellipse = TRUE, groups = data_factors$Differentiation, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  geom_point(aes(col = data_factors$Differentiation, shape = data_factors$Series), size = 2) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "After correction (Ratio A)", color = "Group", shape = "Year") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

## Differential expression
```{r warning=FALSE, message=FALSE}
X_year <- model.matrix(~ Series, data = data_factors)

fit_y <- lmFit(t(ratio_A_corrected), design = X_year, method = "robust", maxit = 1000)

efit_y <- eBayes(fit_y)

topTable(efit_y, coef = 2)

num_spots <- nrow(data_norm_quantile_max)
full_list_y <- topTable(efit_y, coef = 2, number = num_spots,
                        sort.by = "none")
```

Draw heatmap:
```{r warning=FALSE, message=FALSE}
p_above_y <- full_list_y$adj.P.Val <= 0.05
dif_data_ratio_A_year <- data_norm_quantile_max[p_above_y, ]

heatmaply(dif_data_ratio_A_year, main = "Year 1 and 2", fontsize_row = 1, k_col = NA, dendrogram = "col", scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(low = "lightseagreen", high = "orangered3", midpoint = 15))
```

Draw Volcano plot:
```{r warning=FALSE, message=FALSE}
EnhancedVolcano(full_list_y,
                lab = rownames(full_list_y),
                title = "After correction (Ratio A)",
                subtitle = NULL,
                x = 'logFC',
                y = 'adj.P.Val',
                pCutoff = 0.05,
                FCcutoff = 0.1,
                col = c("lightcyan4","#f57002", "#ee9f02", "#0a9278"),
                legend = c("Not significant","Log2FC","Padj","Padj & Log2FC"),
                legendPosition = "right")
```

# Ratio G
*Ratio G* method of batch correction is ratio-based method scaling the expression values by the geometric mean. Batch effect may be corrected by Ratio G method using "bapred" package. 
## Batch correction
```{r warning=FALSE, message=FALSE}
ratio_G_data <- ratiog(t(data_norm_quantile_max), batch)
ratio_G_corrected <- ratio_G_data$xadj
```

## PCA
```{r warning=FALSE, message=FALSE}
pca_ratio_G <- prcomp(ratio_G_corrected, center = T, scale. = F)

ggbiplot(pca_ratio_G, ellipse = TRUE, groups = data_factors$Series, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "After correction (Ratio G)", color = "Year") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

ggbiplot(pca_ratio_G, ellipse = TRUE, groups = data_factors$Differentiation, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  geom_point(aes(col = data_factors$Differentiation, shape = data_factors$Series), size = 2) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "After correction (Ratio G)", color = "Group", shape = "Year") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

## Differential expression
```{r warning=FALSE, message=FALSE}
X_year <- model.matrix(~ Series, data = data_factors)

fit_y <- lmFit(t(ratio_G_corrected), design = X_year, method = "robust", maxit = 1000)

efit_y <- eBayes(fit_y)

topTable(efit_y, coef = 2)

num_spots <- nrow(data_norm_quantile_max)
full_list_y <- topTable(efit_y, coef = 2, number = num_spots,
                        sort.by = "none")
```

Draw heatmap:
```{r warning=FALSE, message=FALSE}
p_above_y <- full_list_y$adj.P.Val <= 0.05
dif_data_ratio_G_year <- data_norm_quantile_max[p_above_y, ]

heatmaply(dif_data_ratio_G_year, main = "Year 1 and 2", fontsize_row = 1, k_col = NA, dendrogram = "col", scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(low = "lightseagreen", high = "orangered3", midpoint = 15))

```

Draw Volcano plot:
```{r warning=FALSE, message=FALSE}
EnhancedVolcano(full_list_y,
                lab = rownames(full_list_y),
                title = "After correction (Ratio G)",
                subtitle = NULL,
                x = 'logFC',
                y = 'adj.P.Val',
                pCutoff = 0.05,
                FCcutoff = 0.1,
                col = c("lightcyan4","#f57002", "#ee9f02", "#0a9278"),
                legend = c("Not significant","Log2FC","Padj","Padj & Log2FC"),
                legendPosition = "right")
```


# Harman 
*Harman* method is based on PCA. It reduces batch effect and keeps user-defined class effects. Harman batch correction is provided by eponymous package "Harman".

## Batch correction
```{r warning=FALSE, message=FALSE}
expt <- data_factors$Differentiation
batch <- data_factors$Series
data.harman <- harman(data_norm_quantile_max, expt, batch)
data_corrected_harman <- reconstructData(data.harman)
```

## PCA
```{r warning=FALSE, message=FALSE}
pca_harman <- prcomp(t(data_corrected_harman), center = T, scale. = F)

ggbiplot(pca_harman, ellipse = TRUE, groups = data_factors$Series, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "After correction (Harman)", color = "Year") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

ggbiplot(pca_harman, ellipse = TRUE, groups = data_factors$Differentiation, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  geom_point(aes(col = data_factors$Differentiation, shape = data_factors$Series), size = 2) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "After correction (Harman)", color = "Group", shape = "Year") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

## Differential expression
```{r warning=FALSE, message=FALSE}
X_year <- model.matrix(~ Series, data = data_factors)

fit_y <- lmFit(data_corrected_harman, design = X_year, method = "robust", maxit = 1000)

efit_y <- eBayes(fit_y)

topTable(efit_y, coef = 2)

num_spots <- nrow(data_norm_quantile_max)
full_list_y <- topTable(efit_y, coef = 2, number = num_spots,
                        sort.by = "none")
```

Draw heatmap:
```{r warning=FALSE, message=FALSE}
p_above_y <- full_list_y$adj.P.Val <= 0.05
dif_data_harman_year <- data_norm_quantile_max[p_above_y, ]

heatmaply(dif_data_harman_year, main = "Year 1 and 2", fontsize_row = 1, k_col = NA, dendrogram = "col", scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(low = "lightseagreen", high = "orangered3", midpoint = 15))
```

Draw Volcano plot:
```{r warning=FALSE, message=FALSE}
EnhancedVolcano(full_list_y,
                lab = rownames(full_list_y),
                title = "After correction (Harman)",
                subtitle = NULL,
                x = 'logFC',
                y = 'adj.P.Val',
                pCutoff = 0.05,
                FCcutoff = 0.1,
                col = c("lightcyan4","#f57002", "#ee9f02", "#0a9278"),
                legend = c("Not significant","Log2FC","Padj","Padj & Log2FC"),
                legendPosition = "right")
```

# Combat
## Batch correction
```{r warning=FALSE, message=FALSE}
pheno <- data_factors
edata <- data_norm_quantile_max
colnames(pheno)[5] <- "batch"

mod <- model.matrix(~ Differentiation + Health, data = pheno)
mod0 <- model.matrix(~ 1, data = pheno)

n.sv <- num.sv(edata, mod, method = "leek")
n.sv

svobj <- sva(edata, mod, mod0, n.sv = n.sv)

modSv <- cbind(mod, svobj$sv)
mod0Sv <- cbind(mod0, svobj$sv)
pValuesSv <- f.pvalue(edata, modSv, mod0Sv)
qValuesSv <- p.adjust(pValuesSv,method = "BH")

fit <- lmFit(edata,modSv)

contrast.matrix <- cbind("C1"= c(-1, 1, 0, rep(0,svobj$n.sv)),"C2"= c(0, -1, 1, rep(0,svobj$n.sv)))
fitContrasts <- contrasts.fit(fit,contrast.matrix)

eb <- eBayes(fitContrasts)

batch <- pheno$batch

modcombat <- model.matrix(~ 1, data = pheno)
combat_edata <- ComBat(dat = edata, batch = batch, mod = modcombat, par.prior = TRUE, prior.plots = TRUE)
```

## PCA
```{r warning=FALSE, message=FALSE}
pca_sva_combat <- prcomp(t(combat_edata), center = T, scale. = F)

ggbiplot(pca_sva_combat, ellipse = TRUE, groups = data_factors$Series, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "After correction (ComBat)", color = "Year") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

ggbiplot(pca_sva_combat, ellipse = TRUE, groups = data_factors$Differentiation, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  geom_point(aes(col = data_factors$Differentiation, shape = data_factors$Series), size = 2) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "After correction (ComBat)", color = "Group", shape = "Year") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

## Differential expression
```{r warning=FALSE, message=FALSE}
X <- model.matrix(~ Series, data = data_factors)

fit <- lmFit(combat_edata, design = X, method = "robust", maxit = 1000)

efit <- eBayes(fit)

topTable(efit, coef = 2)

num_spots <- nrow(combat_edata)
full_list <- topTable(efit, coef = 2, number = num_spots,
                      sort.by = "none")
```

Draw heatmap:
```{r warning=FALSE, message=FALSE}
p_above <- full_list$adj.P.Val <= 0.05
dif_data_combat <- combat_edata[p_above, ]

heatmaply(dif_data_combat, main = "Control vs Healthy", fontsize_row = 1, k_col = 3, dendrogram = "col", scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(low = "lightseagreen", high = "orangered3", midpoint = 15))
```

Draw Volcano plot:
```{r warning=FALSE, message=FALSE}
EnhancedVolcano(full_list,
                lab = rownames(full_list),
                title = "After correction (ComBat)",
                subtitle = NULL,
                x = 'logFC',
                y = 'adj.P.Val',
                pCutoff = 0.05,
                FCcutoff = 0.1,
                col = c("lightcyan4","#f57002", "#ee9f02", "#0a9278"),
                legend = c("Not significant","Log2FC","Padj","Padj & Log2FC"),
                legendPosition = "right")
```

# Comparison of batch correction methods
Firstly, we compared the number of remaining proteins whose expression differ depending on year of experiment. Ideally, there must be no such proteins.
```{r warning=FALSE, message=FALSE}
x <- list(rownames(dif_data_limma), rownames(combat_edata), rownames(dif_data_ratio_A_year), rownames(dif_data_ratio_G_year), rownames(dif_data_harman_year))

ggVennDiagram(x,category.names = c("BMC","Combat", "Ratio A", "Ratio G", "Harman"), label_alpha = 0) +
  scale_fill_gradient(low = "palegreen3", high = "#0a9278") +
  labs(title = "Differentially expressed proteins after \n batch correction \n (condition = 'Year')") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

It could be seen that ComBat method considerably reduced the number of such proteins (only 2 proteins remained). Interestingly, Ratio G method decreased the number of undesirable proteins, but most of them (59) were unique.

# PLS-DA 
Secondly, we compared PLS-DA results. PLS-DA (Partial least squares-discriminant analysis) is a linear classification model that is able to predict the class of samples. Here we used PLS-DA in order to compare prediction of classes:

* Control cells obtained in 1 year
* Control cells obtained in 2 year
* Differentiated cells obtained in 1 year 
* Differentiated cells obtained in 2 year

Classes of one year should be predicted with similar probability

## Data without correction
```{r warning=FALSE, message=FALSE}
data_factors$Series_Differentiation <- factor(apply(data_factors[, c(5, 4)], 1, paste, collapse = "_"))

X <- t(data_norm_quantile_max)
Y <- data_factors$Series_Differentiation 
summary(Y)

list.keepX <- c(5:10,  seq(20, 100, 10))
set.seed(30) 
tune.splsda.srbct <- tune.splsda(X, Y, ncomp = 3,
                                 validation = 'Mfold',
                                 folds = 3, dist = 'max.dist', progressBar = FALSE,
                                 measure = "BER", test.keepX = list.keepX,
                                 nrepeat = 10)   

ncomp <- tune.splsda.srbct$choice.ncomp$ncomp 
ncomp

select.keepX <- tune.splsda.srbct$choice.keepX[1:ncomp] 
select.keepX


MyResult.splsda.fixed <- splsda(X, Y,ncomp = ncomp, keepX = select.keepX)

layout(matrix(c(1, 2, 3, 3, 3, 3), 2, 3))
plotLoadings(MyResult.splsda.fixed, comp = 1, size.name = 1, size.title = 1.2, title = "Loadings\n on 1st component", contrib = "max", legend = FALSE, col.ties="black", ndisplay = 10)
plotLoadings(MyResult.splsda.fixed, comp = 2, size.name = 1, size.title = 1.2, title = "Loadings\n on 2nd component", contrib = "max",ndisplay = 10,  legend = FALSE, col.ties="black")
plotIndiv(MyResult.splsda.fixed, ind.names = F, ellipse = T, style = "graphics", abline = TRUE, cex = 2, pch = 19, size.axis = 1.2, size.xlabel = 1.5, size.ylabel = 1.5, title = "sPLS-DA ordination of samples (No correction)", size.title = 1.5)
legend("bottomright", legend = levels(data_factors$Series_Differentiation), cex = 1, fill = color.mixo(1:4), bty = "n")
```

```{r}
auc.plsda <- auroc(MyResult.splsda.fixed)
```

## BMC
```{r include=FALSE}
dev.off()
```
```{r}
data_factors$Series_Differentiation <- factor(apply(data_factors[, c(5, 4)], 1, paste, collapse = "_"))

X <- t(data_limma_corrected)
Y <- data_factors$Series_Differentiation
summary(Y)


list.keepX <- c(5:10,  seq(20, 100, 10))
set.seed(30) 
tune.splsda.srbct <- tune.splsda(X, Y, ncomp = 3,
                                 validation = 'Mfold',
                                 folds = 3, dist = 'max.dist', progressBar = FALSE,
                                 measure = "BER", test.keepX = list.keepX,
                                 nrepeat = 10)   

ncomp <- tune.splsda.srbct$choice.ncomp$ncomp 
ncomp

select.keepX <- tune.splsda.srbct$choice.keepX[1:ncomp] 
select.keepX


MyResult.splsda.fixed <- splsda(X, Y,ncomp = ncomp, keepX = select.keepX)

layout(matrix(c(1, 2, 3, 3, 3, 3), 2, 3))
plotLoadings(MyResult.splsda.fixed, comp = 1, size.name = 1, size.title = 1.2, title = "Loadings\n on 1st component", contrib = "max", legend = FALSE, col.ties="black", ndisplay = 10)
plotLoadings(MyResult.splsda.fixed, comp = 2, size.name = 1, size.title = 1.2, title = "Loadings\n on 2nd component", contrib = "max",ndisplay = 10,  legend = FALSE, col.ties="black")
plotIndiv(MyResult.splsda.fixed, ind.names = F, ellipse = T, style = "graphics", abline = TRUE, cex = 2, pch = 19, size.axis = 1.2, size.xlabel = 1.5, size.ylabel = 1.5, title = "sPLS-DA ordination of samples (BMC)", size.title = 1.5)
legend(x = "topleft", legend = levels(data_factors$Series_Differentiation), cex = 1, fill = color.mixo(1:4), bty = "n")
```

```{r}
auc.plsda <- auroc(MyResult.splsda.fixed)
```

## Ratio A
```{r include=FALSE}
dev.off()
```

```{r}
data_factors$Series_Differentiation <- factor(apply(data_factors[, c(5, 4)], 1, paste, collapse = "_"))

X <- ratio_A_corrected
Y <- data_factors$Series_Differentiation
summary(Y)


list.keepX <- c(5:10,  seq(20, 100, 10))
set.seed(30) 
tune.splsda.srbct <- tune.splsda(X, Y, ncomp = 3,
                                 validation = 'Mfold',
                                 folds = 3, dist = 'max.dist', 
                                 progressBar = FALSE,
                                 measure = "BER",
                                 test.keepX = list.keepX,
                                 nrepeat = 10)   

ncomp <- tune.splsda.srbct$choice.ncomp$ncomp 
ncomp

select.keepX <- tune.splsda.srbct$choice.keepX[1:ncomp] 
select.keepX


MyResult.splsda.fixed <- splsda(X, Y,ncomp = ncomp, keepX = select.keepX)

layout(matrix(c(1, 2, 3, 3, 3, 3), 2, 3))
plotLoadings(MyResult.splsda.fixed, comp = 1, size.name = 1, size.title = 1.2, title = "Loadings\n on 1st component", contrib = "max", legend = FALSE, col.ties="black", ndisplay = 10)
plotLoadings(MyResult.splsda.fixed, comp = 2, size.name = 1, size.title = 1.2, title = "Loadings\n on 2nd component", contrib = "max",ndisplay = 10,  legend = FALSE, col.ties="black")
plotIndiv(MyResult.splsda.fixed, ind.names = F, ellipse = T, style = "graphics", abline = TRUE, cex = 2, pch = 19, size.axis = 1.2, size.xlabel = 1.5, size.ylabel = 1.5, title = "sPLS-DA ordination of samples (Ratio A)", size.title = 1.5)
legend(x = "topleft", legend = levels(data_factors$Series_Differentiation), cex = 1, fill = color.mixo(1:4), bty = "n")
```
```{r}
auc.plsda <- auroc(MyResult.splsda.fixed)
```

## Ratio G
```{r include=FALSE}
dev.off()
```
```{r}
data_factors$Series_Differentiation <- factor(apply(data_factors[, c(5, 4)], 1, paste, collapse = "_"))

X <- ratio_G_corrected
Y <- data_factors$Series_Differentiation

list.keepX <- c(5:10,  seq(20, 100, 10))
set.seed(30) 
tune.splsda.srbct <- tune.splsda(X, Y, ncomp = 3,
                                 validation = 'Mfold',
                                 folds = 3, dist = 'max.dist', 
                                 progressBar = FALSE,
                                 measure = "BER",
                                 test.keepX = list.keepX,
                                 nrepeat = 10)   

ncomp <- tune.splsda.srbct$choice.ncomp$ncomp 
ncomp

select.keepX <- tune.splsda.srbct$choice.keepX[1:ncomp] 
select.keepX
```

```{r}
MyResult.splsda.fixed <- splsda(X, Y,ncomp = ncomp, keepX = select.keepX)

layout(matrix(c(1, 2, 3, 3, 3, 3), 2, 3))
plotLoadings(MyResult.splsda.fixed, comp = 1, size.name = 1, size.title = 1.2, title = "Loadings\n on 1st component", contrib = "max", legend = FALSE, col.ties="black", ndisplay = 10)
plotLoadings(MyResult.splsda.fixed, comp = 2, size.name = 1, size.title = 1.2, title = "Loadings\n on 2nd component", contrib = "max",ndisplay = 10,  legend = FALSE, col.ties="black")
plotIndiv(MyResult.splsda.fixed, ind.names = F, ellipse = T, style = "graphics", abline = TRUE, cex = 2, pch = 19, size.axis = 1.2, size.xlabel = 1.5, size.ylabel = 1.5, title = "sPLS-DA ordination of samples (Ratio G)", size.title = 1.5)
legend(x = "topleft", legend = levels(data_factors$Series_Differentiation), cex = 1, fill = color.mixo(1:4), bty = "n")
```
```{r}
auc.plsda <- auroc(MyResult.splsda.fixed)
```

## Harman
```{r include=FALSE}
dev.off()
```
```{r}
data_factors$Series_Differentiation <- factor(apply(data_factors[, c(5, 4)], 1, paste, collapse = "_"))

X <- t(data_corrected_harman)
Y <- data_factors$Series_Differentiation

list.keepX <- c(5:10,  seq(20, 100, 10))
set.seed(30) 
tune.splsda.srbct <- tune.splsda(X, Y, ncomp = 3,
                                 validation = 'Mfold',
                                 folds = 3, dist = 'max.dist', progressBar = FALSE,
                                 measure = "BER", test.keepX = list.keepX,
                                 nrepeat = 10)   

ncomp <- tune.splsda.srbct$choice.ncomp$ncomp 
ncomp

select.keepX <- tune.splsda.srbct$choice.keepX[1:ncomp] 
select.keepX


MyResult.splsda.fixed <- splsda(X, Y,ncomp = ncomp, keepX = select.keepX)

layout(matrix(c(1, 2, 3, 3, 3, 3), 2, 3))
plotLoadings(MyResult.splsda.fixed, comp = 1, size.name = 1, size.title = 1.2, title = "Loadings\n on 1st component", contrib = "max", legend = FALSE, col.ties="black", ndisplay = 10)
plotLoadings(MyResult.splsda.fixed, comp = 2, size.name = 1, size.title = 1.2, title = "Loadings\n on 2nd component", contrib = "max",ndisplay = 10,  legend = FALSE, col.ties="black")
plotIndiv(MyResult.splsda.fixed, ind.names = F, ellipse = T, style = "graphics", abline = TRUE, cex = 2, pch = 19, size.axis = 1.2, size.xlabel = 1.5, size.ylabel = 1.5, title = "sPLS-DA ordination of samples (Harman)", size.title = 1.5)
legend(x = "topleft", legend = levels(data_factors$Series_Differentiation), cex = 1, fill = color.mixo(1:4), bty = "n")
```
```{r}
auc.plsda <- auroc(MyResult.splsda.fixed)
```

## Combat
```{r include=FALSE}
dev.off()
```
```{r}
data_factors$Series_Differentiation <- factor(apply(data_factors[, c(5, 4)], 1, paste, collapse = "_"))

X <- t(combat_edata)
Y <- data_factors$Series_Differentiation
summary(Y)


list.keepX <- c(5:10,  seq(20, 100, 10))
set.seed(30) 
tune.splsda.srbct <- tune.splsda(X, Y, ncomp = 3,
                                 validation = 'Mfold',
                                 folds = 3, dist = 'max.dist', 
                                 progressBar = FALSE,
                                 measure = "BER",
                                 test.keepX = list.keepX,
                                 nrepeat = 10)   

ncomp <- tune.splsda.srbct$choice.ncomp$ncomp 
ncomp

select.keepX <- tune.splsda.srbct$choice.keepX[1:ncomp] 
select.keepX
```
```{r}
MyResult.splsda.fixed <- splsda(X, Y,ncomp = ncomp, keepX = select.keepX)

layout(matrix(c(1, 2, 3, 3, 3, 3), 2, 3))
plotLoadings(MyResult.splsda.fixed, comp = 1, size.name = 1, size.title = 1.2, title = "Loadings\n on 1st component", contrib = "max", legend = FALSE, col.ties="black", ndisplay = 10)
plotLoadings(MyResult.splsda.fixed, comp = 2, size.name = 1, size.title = 1.2, title = "Loadings\n on 2nd component", contrib = "max",ndisplay = 10,  legend = FALSE, col.ties="black")
plotIndiv(MyResult.splsda.fixed, ind.names = F, ellipse = T, style = "graphics", abline = TRUE, cex = 2, pch = 19, size.axis = 1.2, size.xlabel = 1.5, size.ylabel = 1.5, title = "sPLS-DA ordination of samples (ComBat)", size.title = 1.5)
legend(x = "topleft", legend = levels(data_factors$Series_Differentiation), cex = 1, fill = color.mixo(1:4), bty = "n")
```
```{r}
auc.plsda <- auroc(MyResult.splsda.fixed)
```

Only application of ComBat resulted in similar prediction probability for classes of first and second years.
Usage of other methods led to higher prediction probability for Classes of first year than for second year.

# gPCA
gPCA (guided PCA) is used to test whether a batch effect exists. "Delta" is a metric that varies between 0 and 1. 1 means that there are batch effect, 0 - no batch. p-value indicates the significance of delta value.

```{r include=FALSE}
# Function for gPCA analysiss may be found in "Functions" directory
source("gPCA.batchdetect.R")
```

```{r}
real <- gPCA.batchdetect(t(data_norm_quantile_max), data_factors$Series)
sva <- gPCA.batchdetect(t(combat_edata), data_factors$Series, nperm = 10000)
limma <- gPCA.batchdetect(t(data_limma_corrected), data_factors$Series)
ratio_A <- gPCA.batchdetect(ratio_A_corrected, data_factors$Series)
ratio_G <- gPCA.batchdetect(ratio_G_corrected, data_factors$Series)
harman <- gPCA.batchdetect(t(data_corrected_harman), data_factors$Series)

gPCA_table <- data.frame("Correction method" = c("No correction",
                                                 "ComBat",
                                                 "BMC", 
                                                 "Ratio A", 
                                                 "Ratio G", 
                                                 "Harman"),
                         "gPCA delta" = c(real$delta,
                                          sva$delta,
                                          limma$delta,
                                          ratio_A$delta,
                                          ratio_G$delta,
                                          harman$delta), 
                         "P-value" = c(real$p.val,
                                       sva$p.val,
                                       limma$p.val,
                                       ratio_A$p.val,
                                       ratio_G$p.val,
                                       harman$p.val))
gPCA_table
```


# Conclusion
Thus, **ComBat** method was used in further differential expression and GO enrichment analyses.

