---
title: "Data Preparation, Investigation, Differentail expression"
author: "Katerina Danko"
output:
    html_document:
      code_folding: show
      theme: united
      toc: true
      toc_depth: 3
      toc_float: true
      number_section: true
---

```{css, echo=FALSE}
.header-section-number::after {
  content: ".";
}
```

List of used packages:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

```

```{r echo = FALSE, message=FALSE, warning=FALSE}
# Missing packages must be installed for proper compilation 
if (!require(limma)) BiocManager::install("limma")
if (!require(devtools)) install.packages("devtools")
if (!require(readxl)) install.packages("readxl")
if (!require(readr)) install.packages("readr")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(impute)) BiocManager::install("impute")
if (!require(ggbiplot)) install_github("vqv/ggbiplot")
if (!require(reactable)) install.packages("reactable")
if (!require(heatmaply)) BiocManager::install("heatmaply")
if (!require(EnhancedVolcano)) BiocManager::install("EnhancedVolcano")
```


```{r echo = T, warning=FALSE, message=FALSE, eval=TRUE}
library(limma)
library(readxl)
library(readr)
library(impute)
library(ggbiplot)
library(reactable)
library(heatmaply)
library(EnhancedVolcano)
library(ggVennDiagram)
```

```{r echo = FALSE, warning=FALSE, message=FALSE, eval=TRUE}
theme_set(theme_bw())
```

# Comparison of protemic software
We had two softs for mass-spectrometry results analysis: Peaks Xpro and MaxQuant. 
Firstly, we compared identified proteins using these two softs.
```{r message=FALSE, warning=FALSE}
# All data may be found in "Data" directory
maxq <- read_xlsx("raw_proteomics_MaxQuant_filtered.xlsx")
peaks <- read_csv("data_Peaks.csv")
id_list <- list(peaks$Gene_id, maxq$gene_id)

ggVennDiagram(id_list, category.names = c("Peaks", "MaxQuant"), label_alpha = 0) +
  scale_fill_gradient(low = "palegreen3", high = "#0a9278") +
  labs(title = "Protein identification") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```
We chose data, obtained by using Peaks software as it identified much more proteins 

# Read files and optimize datasets
```{r warning=FALSE, message=FALSE}
# Read raw data obtained from "Peaks" software
data <- read.csv("data_Peaks.csv", stringsAsFactors = F)

# Make tidy data
data$Accession_id <- unlist(lapply(data$Accession, function(x) unlist(strsplit(x, "\\|")[1])[1]))
data$temp <- unlist(lapply(data$Accession, function(x) unlist(strsplit(x, "\\|")[1])[2]))
data$Protein_name <- unlist(lapply(data$temp, function(x) unlist(strsplit(x, "_")[1])[1]))
data <- data[, -111]


# Read factor table
data_factors <- read_xlsx("factor_Peaks.xlsx")
data_factors$Health <- as.factor(data_factors$Health)
data_factors$Differentiation <- as.factor(data_factors$Differentiation)
data_factors$Series <- as.factor(data_factors$Series)


colnames(data)[39:71] <- data_factors$sample

reactable(data[, c(111, 39:71)])

reactable(data_factors)
```

# Dealing with NA
Proteomics data may possess NA values which should be treated carefully. Here we tried two approaches:      

* Drop columns where the number of NA is more than 2 (MAX drop)
* Drop columns where the number of NA is more than a half of observations (MIN drop)
```{r warning=FALSE, message=FALSE}
t_data <- t(data[, 39:71])

colnames(t_data) <- data$Protein_name

## Remove columns where the number of NA is greater than half of observations
cond_1 <- colSums(is.na(t_data)) <=  nrow(t_data) / 2
t_data_mindrop <- t_data[, cond_1]
sum(colSums(is.na(t_data)) >=  nrow(t_data) / 2)

## Remove columns where the number of NA is greater than 2
cond_2 <- colSums(is.na(t_data)) <  2
t_data_maxdrop <- t_data[, cond_2]
sum(colSums(is.na(t_data)) >  2)

## Imputation of NA 
data_maxdrop <- t(impute.knn(t_data_maxdrop, k = 10)$data)
sum(is.na(data_maxdrop))

data_mindrop <- t(impute.knn(t_data_mindrop, k = 10)$data)
sum(is.na(data_mindrop))
```

Application of MAX drop approach resulted in considerable reduction of proteins number (approximately two-thirds of all proteins), whereas usage MIN drop led to more slight decrease in protein number (around one third of all proteins).

# Data Normalization
Log2 transformation and quantile transformation were used as it had demonstrated the best normalization results.
```{r}
# MAX DROP (Drop NA if > 2)
data_norm_log_max <- log2(data_maxdrop + 1)
data_norm_quantile_max <- normalizeQuantiles(data_norm_log_max)
boxplot(data_norm_quantile_max,col = "#0a9278",
        border = "black",
        main = "Log2 + Quantile normalization",
        ylab = "intensities", 
        xlab = "samples")

# MIN DROP (Drop NA if >  half of observations)
data_norm_log_min <- log2(data_mindrop + 1)
data_norm_quantile_min <- normalizeQuantiles(data_norm_log_min)
boxplot(data_norm_quantile_min, col = "#0a9278", 
        border = "black",
        main = "Log2 + Quantile normalization",
        ylab = "intensities", 
        xlab = "samples")
```

# PCA
We used PCA to decide which data would be analyzed: MAX drop or MIN drop.

## "MAX drop"
```{r warning=FALSE, message=FALSE}
# Year
data_pca_maxdrop <- prcomp(t(data_norm_quantile_max), center = T, scale. = F)
ggbiplot(data_pca_maxdrop, ellipse = TRUE, groups = data_factors$Series, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "Batch effect in data MAX drop", color = "Year") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

```{r warning=FALSE, message=FALSE}
# Health
ggbiplot(data_pca_maxdrop, ellipse = TRUE, groups = data_factors$Health, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "Data classes (Control vs Health)", color = "Group") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

```{r warning=FALSE, message=FALSE}
# Differentiation
ggbiplot(data_pca_maxdrop, ellipse = TRUE, groups = data_factors$Differentiation, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  geom_point(aes(col = data_factors$Differentiation)) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "Data classes (Control vs Differentiation)", color = "Group") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

## "MIN drop"
```{r warning=FALSE, message=FALSE}
# Year
data_pca_mindrop <- prcomp(t(data_norm_quantile_min), center = T, scale. = F)
ggbiplot(data_pca_mindrop, ellipse = TRUE, groups = data_factors$Series, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "Batch effect in data (MIN drop)", color = "Year") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

```{r warning=FALSE, message=FALSE}
# Health
ggbiplot(data_pca_mindrop, ellipse = TRUE, groups = data_factors$Health, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "Data classes (Control vs Health)", color = "Group") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

```{r warning=FALSE, message=FALSE}
# Differentiation
ggbiplot(data_pca_mindrop, ellipse = TRUE, groups = data_factors$Differentiation, labels = NULL, var.axes = FALSE, alpha = 0.7) +
  geom_point(aes(col = data_factors$Differentiation)) +
  scale_color_manual(values = c("#0a9278","#f57002")) +
  labs(title = "Data classes (Control vs Differentiation)", color = "Group") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

We chose "MAX drop" data for further analysis, because Control and Differentiation classes were more separated from each other than in case of "MIN drop" data.

# Differential expression
We have performed differential expression analysis depending on year of experiment in order to somehow estimate the scale of batch effect. 
```{r warning=FALSE, message=FALSE}
X <- model.matrix(~ Series, data = data_factors)

# Build a linear model for each protein
fit <- lmFit(data_norm_quantile_max, design = X, method = "robust", maxit = 1000)

# Empirical Bayes statistics
efit <- eBayes(fit)

topTable(efit, coef = 2)

num_spots <- nrow(data_norm_quantile_max)
full_list <- topTable(efit, coef = 2, number = num_spots,
                      sort.by = "none")


p_above <- full_list$adj.P.Val <= 0.05
dif_data_year_wo_correction <- data_norm_quantile_max[p_above, ]
sum(p_above)
```

As you may see, there are 707 "differentially expressed" proteins depending on year of experiment.

Draw heatmap:
```{r}
heatmaply(dif_data_year_wo_correction, main = "1 Year vs 2 Year", fontsize_row = 1, dendrogram = "col",  scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(low = "lightseagreen", high = "orangered3", midpoint = 15))
```

Draw Volcano plot:
```{r warning=FALSE, message=FALSE}
EnhancedVolcano(full_list,
                lab = rownames(full_list),
                title = "Batch effect connecting with year\nof experiment",
                subtitle = NULL,
                x = "logFC",
                y = "adj.P.Val",
                pCutoff = 0.05,
                FCcutoff = 0.1,
                legend = c("Not significant","Log2FC","Padj","Padj & Log2FC"),
                legendPosition = "right",
                col = c("lightcyan4","#f57002", "#ee9f02", "#0a9278"))
```