---
title: "Differential expression and GO enrichment analyzes"
author: "Katerina Danko"
output:
    html_document:
      code_folding: show
      theme: united
      toc: true
      toc_depth: 3
      toc_float: true
      number_section: true
---

```{css, echo=FALSE}
.header-section-number::after {
  content: ".";
}
```

List of used packages:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

```

```{r include=FALSE}
# "data.Rdata" and "combat.Rdata" may be found in "Rdata" directory 
load("combat.Rdata")
load("data.Rdata")
```


```{r echo = FALSE, message=FALSE, warning=FALSE}
# Missing packages must be installed for proper compilation 
if (!require(limma)) BiocManager::install("limma")
if (!require(dplyr)) install.packages("dplyr")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(sva)) BiocManager::install("sva")
if (!require(heatmaply)) BiocManager::install("heatmaply")
if (!require(EnhancedVolcano)) BiocManager::install(EnhancedVolcano)
if (!require(clusterProfiler)) BiocManager::install(clusterProfiler)
if (!require(enrichplot)) BiocManager::install(enrichplot)
if (!require(DOSE)) BiocManager::install(DOSE)
if (!require(gprofiler2)) BiocManager::install(gprofiler2)
```

```{r echo = T, warning=FALSE, message=FALSE, eval=TRUE}
library(limma)
library(ggplot2)
library(sva)
library(heatmaply)
library(EnhancedVolcano)
library(dplyr)
library(clusterProfiler)
library(enrichplot)
library(DOSE)
library(gprofiler2)
```

We have chosen ComBat method of batch correction provided by `sva` package.
Two proteins whose expression differed depending on the year of experiment were dropped.
```{r warning=FALSE, message=FALSE}
dropped_genes <- rownames(dif_data_sva)
new_data_combat <- combat_edata[!(rownames(combat_edata) %in% dropped_genes), ]
```

# Diffential expression depending on stage of differentiation
This study aimed to deаine molecular mechanisms of cells osteogenic differentiation. So that we performed differential expression analysis of cells in undifferentiated and in osteogenic differentiation stages.

```{r warning=FALSE, message=FALSE}

# Dif expression
X_diff <- model.matrix(~ Differentiation, data = data_factors)

# Build a linear model for each protein
fit_y <- lmFit(new_data_combat, design = X_diff, method = "robust", maxit = 1000)

# Empirical Bayes statistics
efit_y <- eBayes(fit_y)

topTable(efit_y, coef = 2)

num_spots <- nrow(data_norm_quantile_max)
full_list_diff <- topTable(efit_y, coef = 2, number = num_spots,
                        sort.by = "none")
```

Draw heatmap:
```{r warning=FALSE, message=FALSE}
p_above_y <- full_list_diff$adj.P.Val < 0.05
dif_data_diff <- data_norm_quantile_max[p_above_y, ]
sum(p_above_y)

heatmaply(dif_data_diff, main = "Differentiation vs Control", fontsize_row = 1, dendrogram = "col", scale_fill_gradient_fun = ggplot2::scale_fill_gradient2(low = "lightseagreen", high = "orangered3", midpoint = 15))
```

Draw Volcano plot:
```{r warning=FALSE, message=FALSE}
EnhancedVolcano(full_list_diff,
                lab = rownames(full_list_diff),
                x = 'logFC',
                y = 'adj.P.Val',
                title = "Differentially expressed genes\n condition = 'Differentiation'",
                subtitle = NULL,
                pCutoff = 0.05,
                FCcutoff = 0.1,
                legend = c("Not significant","Log2FC","Padj","Padj & Log2FC"),
                legendPosition = "right",
                col=c("lightcyan4","#f57002", "#ee9f02", "#0a9278"))
```

# GO enrichment analysis
In order to group significantly differentially expressed proteins we have used GO enrichment analysis.

```{r warning=FALSE, message=FALSE}
# keep only the significant proteins results
sig <- subset(full_list_diff, adj.P.Val < 0.05)

# get the significant up-regulated proteins
up <- subset(full_list_diff, logFC > 0)

# get the significant down-regulated proteins
down <- subset(full_list_diff, logFC < 0)
```

```{r warning=FALSE, message=FALSE}
# needed to convert to enrichResult object
up_names <- gconvert(row.names(up))
down_names <- gconvert(row.names(down))
```

## Up-regulated proteins
```{r warning=FALSE, message=FALSE}
# enrichment analysis using proteins names
multi_gp_up_reg <- gost(list("up-regulated" = up_names$name), multi_query = FALSE, evcodes =TRUE)

# modify the g:Profiler data frame
gp_mod_up = multi_gp_up_reg$result[, c("query", "source", "term_id","term_name", "p_value", "query_size", "intersection_size", "term_size", "effective_domain_size", "intersection")]

gp_mod_up <- gp_mod_up[order(gp_mod_up$p_value), ]
gp_mod_up_BP <- gp_mod_up[gp_mod_up$source == "GO:BP", ]

gp_mod_up_BP$GeneRatio <- paste0(gp_mod_up_BP$intersection_size,  "/", gp_mod_up_BP$query_size)
gp_mod_up_BP$BgRatio <- paste0(gp_mod_up_BP$term_size, "/", gp_mod_up_BP$effective_domain_size)

names(gp_mod_up_BP) <- c("Cluster", "Category", "ID", "Description", "p.adjust", "query_size", "Count", "term_size", "effective_domain_size", "geneID", "GeneRatio", "BgRatio")

gp_mod_up_BP$geneID <- gsub(",", "/", gp_mod_up_BP$geneID)
row.names(gp_mod_up_BP) <- gp_mod_up_BP$ID

gp_mod_enrich_up_BP <- new("enrichResult", result = gp_mod_up_BP)
```

Draw enrichment plot:
```{r warning=FALSE, message=FALSE}
enrichplot::dotplot(gp_mod_enrich_up_BP, showCategory = 10) + ggplot2::labs(title = "up-regulated") + ggplot2::scale_color_gradient(low = "lightseagreen", high = "darkorange1")
```

## Down-regulated proteins
```{r warning=FALSE, message=FALSE}
# enrichment analysis using gene names
multi_gp_down_reg <- gost(list("down-regulated" = down_names$name), multi_query = FALSE, evcodes =TRUE)


# modify the g:Profiler data frame
gp_mod_down = multi_gp_down_reg$result[, c("query", "source", "term_id","term_name", "p_value", "query_size", "intersection_size", "term_size", "effective_domain_size", "intersection")]

gp_mod_down <- gp_mod_down[order(gp_mod_down$p_value), ]

# BP
gp_mod_down_BP <- gp_mod_down[gp_mod_down$source == "GO:BP", ]

gp_mod_down_BP$GeneRatio <- paste0(gp_mod_down_BP$intersection_size,  "/", gp_mod_down_BP$query_size)
gp_mod_down_BP$BgRatio <-  paste0(gp_mod_down_BP$term_size, "/", gp_mod_down_BP$effective_domain_size)

names(gp_mod_down_BP) <- c("Cluster", "Category", "ID", "Description", "p.adjust", "query_size", "Count", "term_size", "effective_domain_size", "geneID", "GeneRatio", "BgRatio")

gp_mod_down_BP$geneID <- gsub(",", "/", gp_mod_down_BP$geneID)

gp_mod_enrich_down_BP <- new("enrichResult", result = gp_mod_down_BP)
```

Draw enrichment plot:
```{r warning=FALSE, message=FALSE}
enrichplot::dotplot(gp_mod_enrich_down_BP, showCategory = 10) + ggplot2::labs(title = "down-regulated") + ggplot2::scale_color_gradient(low = "lightseagreen", high = "darkorange1")
```

# Conclusions

* There were considerably more proteins identified by Peaks Xpro than by MaxQuant
* Wrong experimental design that included two-step data analysis in two years led in severe technical batch effect
* Differential expression analysis depending on year of experiment showed 707 “differentially expressed” proteins 
* 5 methods of batch correction were applied: ComBat, BMC, Ratio A, Ratio G, Harman
* Comparison of batch correction methods established that the optimal method is ComBat
* GO enrichment analysis conducted on corrected data revealed that proteins connected with immune response activation were up-regulated in differentiated cells, while proteins participating in cell transporting processes were down-regulated 
